---
title: "Single-Cell RNA-seq Analysis of FSHD (GSE122873)"
author: "Manisha Kalsan"
date: "2026-02-28"
output: html_document
---

```{r setup, include=FALSE}

library(Seurat)
library(Matrix)

library(clusterProfiler)
library(org.Hs.eg.db)

library(biomaRt)
library(ggplot2)
library(DESeq2)
library(apeglm)
library(dplyr)

library(SingleR)
library(celldex)
```

## Single-Cell RNA-seq Analysis of FSHD (GSE122873)


```{r}

files <- list.files(pattern="GSM.*\\.txt$")

# Function to read one file
load_sample <- function(file) {
       # remove txt extension
       sample_name <- gsub(".txt","",file)
       # read the file
       mat <- read.table(file,
                         header=TRUE,
                         row.names=1,
                         sep="\t",
                         check.names=FALSE)
       # convert to matrix
       mat <- Matrix(as.matrix(mat), sparse=TRUE)
       
       # create seurat object
       seu <- CreateSeuratObject(mat,
                                 project=sample_name,
                                 min.cells=3,
                                 min.features=200)
       # assign sample name
       seu$sample_id <- sample_name
       
       return(seu)
}

# Apply to all files
seurat_list <- lapply(files, load_sample)



```

## Preliminary inspection of the datasets

This includes basic inspection data to check the number of genes and cells in each sample, average number of genes across samples.

If the overlap of genes across samples is around 12k or more (12-14K), it is fairly good data.
If the common genes are <10K, the overlap is low.

```{r}


# check the length of genenames before merging
lapply(seurat_list, nrow)

head(rownames(seurat_list[[1]]))

# check length of common genes
length(Reduce(intersect, lapply(seurat_list, rownames)))

# the overlap of genes across all samples is good enough (~12k)

# with varying number of genes in data from each sample, we can either take union 
# of the genes (standard practice) or we can take intersection (common genes; more strict) 

all_seu <- merge(seurat_list[[1]], y=seurat_list[-1])

table(all_seu$sample_id)

dim(all_seu)


```

The data contained in the files is in the format gene vs UMIs where each numerical value is the cell wise gene expression.

The gene ID in the data are at the level of Ensembl IDs, which can create issues, hence, we convert them to gene names first. We use biomart package to first map the Ensembl IDs to HGNC symbols and then use these symbols to assign names to genes in the datasets.

```{r}
mart <- useMart("ensembl", dataset="hsapiens_gene_ensembl")

gene_map <- getBM(attributes=c("ensembl_gene_id","hgnc_symbol"),
                  filters="ensembl_gene_id",
                  values=rownames(all_seu),
                  mart=mart)

gene_symbols <- gene_map$hgnc_symbol
names(gene_symbols) <- gene_map$ensembl_gene_id

new_names <- gene_symbols[rownames(all_seu)]

# print heads to check the naming
head (gene_map)
head(rownames(all_seu))
head(gene_symbols)
head(new_names)

# replace missing HGNC symbols with original Ensembl ID
new_names[is.na(new_names) | new_names == ""] <- 
       rownames(all_seu)[is.na(new_names) | new_names == ""]

# make the names unique (to avoid issues with duplicate rownames)
new_names <- make.unique(new_names)

# replace ensembl IDs with HGNC symbols
rownames(all_seu) <- new_names

head(rownames(all_seu))
```

The gene names may contain version numbers. We remove them by removing version numbers and making the names unique.

```{r}
# check if any of the IDs contain version numbers
any(grepl("\\.", rownames(all_seu)))

# if yes, do this
# rownames(all_seu) <- sub("\\..*", "", rownames(all_seu))

```

## Clustering of cells

We check the seperation of cells in the datasets.

The aim of doing this exercise is to understand how well the cells are seperated. For this analysis, we first normalize the data, find variable features, scale the data and perform PCA or UMAP.
This is done to visualize the cells in two dimensional space.

We try to see that there is mixing of cells at the cluster levels and that the clusters are not directly associated with sample IDs, in which case, the batch effect issue becomes evident.

```{r}
# check clustering in the data
all_seu <- NormalizeData(all_seu)
all_seu <- FindVariableFeatures(all_seu, selection.method="vst", nfeatures=2000)
all_seu <- ScaleData(all_seu)
all_seu <- RunPCA(all_seu)

DimPlot(all_seu, group.by="sample_id")

all_seu <- RunUMAP(all_seu, dims=1:20)
DimPlot(all_seu, reduction = "umap")

```
## Filtering of data

Before moving on to the downstream analysis of data, we first filter the data based on the number of read per cell, number of genes per cell, and mitochondrial content. These filtering criteria are applied to filter out possible doublets, or dying/ apoptotic cells.


```{r}
all_seu[["percent.mt"]] <- PercentageFeatureSet(all_seu, pattern="^MT-")

VlnPlot(all_seu, features=c("nFeature_RNA","nCount_RNA","percent.mt"), group.by="sample_id")

# Filtering of data
# subset the data with thresholds defined on the basis of visual inspection
all_seu <- subset(all_seu, subset = nFeature_RNA >500 & nFeature_RNA < 5000 & percent.mt < 15)

# check again the clustering pattern
all_seu <- NormalizeData(all_seu)
all_seu <- FindVariableFeatures(all_seu, selection.method="vst", nfeatures=2000)
all_seu <- ScaleData(all_seu)
all_seu <- RunPCA(all_seu)

```

Once the data is filtered, we again visualize the data to see heterogeneity of clusters.
We try to see whether there are specific sample dominated clusters.

```{r}
VizDimLoadings(all_seu, dims = 1:2, reduction = "pca")
# DimPlot(all_seu, reduction = "pca") + NoLegend()


DimHeatmap(all_seu, dims = 1:9, cells = 500, balanced = TRUE)
# ElbowPlot(all_seu)

# print weights of genes for PCs
# Loadings(all_seu[["pca"]])[,1]

all_seu <- RunUMAP(all_seu, dims=1:20)
DimPlot(all_seu, reduction = "umap")
```


Now, there were isolated groups in the UMAP plot prior filtering and we visualize that they are still there so it is most likely not because of High mitochondrial cells, Empty droplets, or Extreme low-quality cells.


```{r}
all_seu <- FindNeighbors(all_seu, dims = 1:20)
all_seu <- FindClusters(all_seu, resolution = 0.5)

# look at cluster IDs
head(Idents(all_seu), 5)


all_seu <- RunUMAP(all_seu, dims=1:20)

DimPlot(all_seu, reduction = "umap")
DimPlot(all_seu, group.by="sample_id")
```

## Differential expression

We can either find differentially expressed genes (DEGs) across all clusters, between specific clusters or FSHD vs CTRL samples.

For predicting DEGs, we use FindMarkers. Here, we are identifying DEGs for cluster 0 campared to other clusters, cluster 6 compared to other clusters, and finally, gene biomarkers which are differentially expressed across all cell clusters.


```{r}
all_seu <- JoinLayers(all_seu)

# Finding differentially expressed features (cluster biomarkers)
cluster0.markers <- FindMarkers(all_seu, ident.1 = 0)
head(cluster0.markers, n = 5)

cluster6.markers <- FindMarkers(all_seu, ident.1 = 6)
head(cluster6.markers, n = 5)

# find markers for every cluster compared to all remaining cells, report only the positive
# ones
all_seu.markers <- FindAllMarkers(all_seu, only.pos = TRUE)
all_seu.markers %>%
       group_by(cluster) %>%
       dplyr::filter(avg_log2FC > 1, pct.1 > 0.7, pct.2 < 0.3)
```

Now we assign conditions to FSHD and control samples to find biomarkers across these conditions.

```{r}
# assign conditions
all_seu$type <- ifelse(grepl("CTRL", all_seu$sample_id),
                            "CTRL", ifelse(grepl("FSHD1", all_seu$sample_id),
                                           "FSHD1", "FSHD2"))
all_seu$condition <- ifelse(grepl("CTRL", all_seu$sample_id),
                            "CTRL",
                            "FSHD")

Idents(all_seu) <- "condition"

fshd_ctrl <- FindMarkers(all_seu,
                         group.by="condition",
                         ident.1="FSHD",
                         ident.2="CTRL")
fshd_ctrl <- fshd_ctrl[order(fshd_ctrl$avg_log2FC, decreasing=TRUE), ]
head(fshd_ctrl, 20)
```

As this data is from a published study, we will use the data that the authors used to find FSHD relevant transcription factor DUX4 gene list to find cells which are DUX4 positive and identify those which are DUX4 negative.

```{r}
# now, load DUX4-67 gene set that the author have used 
dux4_genes <- read.table("DUX_gene_set", header = F)

dux4_present <- intersect(dux4_genes[,1], rownames(all_seu))
expr_mat <- GetAssayData(all_seu, layer="counts")[dux4_present, ]

all_seu$DUX4_gene_count <- colSums(expr_mat > 0)
all_seu$DUX4_cum_reads  <- colSums(expr_mat)

all_seu$DUX4_affected <- all_seu$DUX4_gene_count >= 5


ggplot(all_seu@meta.data,
       aes(x = DUX4_gene_count,
           y = DUX4_cum_reads,
           color = sample_id)) +
       geom_point(alpha = 0.9, size=5) +
       geom_vline(xintercept = 5, linetype = "dashed") +
       theme_classic()

# UMAP colored by DUX4 count
FeaturePlot(all_seu,
            features = "DUX4_gene_count",
            cols = c("grey90", "red"))

# Count DUX4 Cells Per Sample 
table(all_seu$sample_id,
      all_seu$DUX4_affected)

# proportion per sample
prop.table(table(all_seu$sample_id,
                 all_seu$DUX4_affected), 1)[,2]

# cluster enrichment
table(all_seu$seurat_clusters,
      all_seu$DUX4_affected)

# subset only FSHD cells
fshd_only <- subset(all_seu, subset = condition == "FSHD")

# set identity
Idents(fshd_only) <- "DUX4_affected"

# differential expression
dux4_vs_rest <- FindMarkers(fshd_only,
                            ident.1 = TRUE,
                            ident.2 = FALSE,
                            min.pct = 0.1,
                            logfc.threshold = 0.25)



```

Another way of finding DEGs is using pseudobul analysis where we aggregate data from single cells to represent bulk level expressions and then perform DE analysis as we do with bulk RNAseq.
We use DESeq2 here to find DEGs. at two levels: FSHD vs control and DUX4 positive vs DUX4 negative cells.

```{r}

# Pseudobulk analysis
counts <- as.matrix(GetAssayData(all_seu, layer = "counts"))

# sample level aggregation
sample_ids <- all_seu$sample_id

# make sure sample_ids are aligned
sample_ids <- all_seu$sample_id
names(sample_ids) <- colnames(counts)

# aggregate
pseudobulk_counts <- do.call(cbind,
                             lapply(unique(sample_ids), function(s) {
                                    Matrix::rowSums(counts[, sample_ids == s])
                             })
)

colnames(pseudobulk_counts) <- unique(sample_ids)

dim(pseudobulk_counts)

```

Perform steps for DESeq2

```{r}
sample_metadata <- data.frame(
       row.names = colnames(pseudobulk_counts),
       condition = ifelse(grepl("CTRL", colnames(pseudobulk_counts)),
                          "CTRL", "FSHD")
)

dds <- DESeqDataSetFromMatrix(
       countData = pseudobulk_counts,
       colData = sample_metadata,
       design = ~ condition
)

dds <- DESeq(dds)
res <- results(dds)

colSums(pseudobulk_counts)

# Right now you are using raw MLE log2FC.
# 
# With small sample size, we should shrink them.
# 
# This makes volcano plots much more stable and paper-like.

res_shrunk <- lfcShrink(dds,
                        coef = "condition_FSHD_vs_CTRL",
                        type = "apeglm")
res_df <- as.data.frame(res_shrunk)
res_df$gene <- rownames(res_df)

# remove NA padj
res_df <- res_df[!is.na(res_df$padj), ]

# compute -log10
res_df$neg_log10_p <- -log10(res_df$padj)

# define significance
res_df$significance <- "NS"
res_df$significance[res_df$log2FoldChange > 1 & res_df$padj < 0.05] <- "Up"
res_df$significance[res_df$log2FoldChange < -1 & res_df$padj < 0.05] <- "Down"

# number of up- and down-reg genes
table(res_df$significance)

```

Visualize the data using a volcano plot.
it is an ideal visualization to show upregulated and down regulated genes.

```{r}
# Make a volcano plot
ggplot(res_df,
       aes(x = log2FoldChange,
           y = neg_log10_p,
           color = significance)) +
       geom_point(alpha = 0.7, size = 2) +
       scale_color_manual(values = c("Up"="red",
                                     "Down"="blue",
                                     "NS"="grey70")) +
       geom_vline(xintercept = c(-1,1), linetype="dashed") +
       geom_hline(yintercept = -log10(0.05), linetype="dashed") +
       theme_classic() +
       labs(title="FSHD vs CTRL (Pseudobulk DESeq2)",
            x="Log2 Fold Change",
            y="-log10(adj p-value)")

```

Now we do the same aalysis for DUX4 positive vs DUX4 negative cells.
```{r}

# DUX4+ vs DUX4 - cell based pseudobulk analysis DEGs
fshd_only <- subset(all_seu, subset = condition == "FSHD")

counts <- GetAssayData(fshd_only, layer = "counts")
counts <- as(counts, "dgCMatrix")
group_id <- paste(fshd_only$sample_id,
                  fshd_only$DUX4_affected,
                  sep = "_")

names(group_id) <- colnames(counts)

pseudobulk_counts <- do.call(cbind,
                             lapply(unique(group_id), function(g) {
                                    Matrix::rowSums(counts[, group_id == g])
                             })
)

colnames(pseudobulk_counts) <- unique(group_id)

dim(pseudobulk_counts)

# create metadata
sample_info <- data.frame(
       row.names = colnames(pseudobulk_counts)
)

sample_info$sample <- sub("_(TRUE|FALSE)$", "", rownames(sample_info))
sample_info$DUX4_status <- ifelse(grepl("TRUE$", rownames(sample_info)),
                                  "DUX4pos",
                                  "DUX4neg")


dds <- DESeqDataSetFromMatrix(
       countData = pseudobulk_counts,
       colData = sample_info,
       design = ~ sample + DUX4_status
)

dds <- DESeq(dds)

res <- results(dds, contrast = c("DUX4_status", "DUX4pos", "DUX4neg"))

res_shrunk <- lfcShrink(dds,
                        coef = "DUX4_status_DUX4pos_vs_DUX4neg",
                        type = "apeglm")


res_df <- as.data.frame(res_shrunk)
res_df$gene <- rownames(res_df)
res_df <- res_df[!is.na(res_df$padj), ]
res_df$neg_log10_p <- -log10(res_df$padj)

res_df$significance <- "NS"
res_df$significance[res_df$log2FoldChange > 1 & res_df$padj < 0.05] <- "Up"
res_df$significance[res_df$log2FoldChange < -1 & res_df$padj < 0.05] <- "Down"

ggplot(res_df,
       aes(x = log2FoldChange,
           y = neg_log10_p,
           color = significance)) +
       geom_point(alpha = 0.7, size = 2) +
       scale_color_manual(values = c("Up"="red",
                                     "Down"="blue",
                                     "NS"="grey70")) +
       geom_vline(xintercept = c(-1,1), linetype="dashed") +
       geom_hline(yintercept = -log10(0.05), linetype="dashed") +
       theme_classic() +
       labs(title="DUX4+ vs DUX4− (Pseudobulk)",
            x="Log2 Fold Change",
            y="-log10(adj p-value)")
```

we clean the data and select the upregulated and downregulated genes on the basis of the adjust p-value and log2FoldChange thresholds of 0.05 and 1 respectively.

```{r}
# remove NA
res_df_clean <- res_df[!is.na(res_df$padj), ]

# create ranked vector
gene_list <- res_df_clean$log2FoldChange
names(gene_list) <- res_df_clean$gene

# define thresholds
up_genes <- res_df_clean$gene[
       res_df_clean$padj < 0.05 & res_df_clean$log2FoldChange > 1
]

down_genes <- res_df_clean$gene[
       res_df_clean$padj < 0.05 & res_df_clean$log2FoldChange < -1
]

length(up_genes)
length(down_genes)
```

## Pathway enrichment analysis
Now, we perform patway enrichment analysis of down- and up-regulated genes for DUX4 positive vs DUX4 negative cells

```{r}
# UP genes
up_entrez <- bitr(up_genes,
                  fromType = "SYMBOL",
                  toType = "ENTREZID",
                  OrgDb = org.Hs.eg.db)

# DOWN genes
down_entrez <- bitr(down_genes,
                    fromType = "SYMBOL",
                    toType = "ENTREZID",
                    OrgDb = org.Hs.eg.db)



kegg_up <- enrichKEGG(
       gene = up_entrez$ENTREZID,
       organism = "hsa",
       pAdjustMethod = "BH"
)

head(kegg_up)

kegg_down <- enrichKEGG(
       gene = down_entrez$ENTREZID,
       organism = "hsa",
       pAdjustMethod = "BH"
)

head(kegg_down)

p_up <- dotplot(kegg_up, showCategory = 15) +
       ggtitle("KEGG Pathways – Upregulated")

p_down <- dotplot(kegg_down, showCategory = 15) +
       ggtitle("KEGG Pathways – Downregulated")

p_up+p_down

```

## Cell type annotation

Now we will assign biological identities to clusters.

```{r}

ref <- HumanPrimaryCellAtlasData()

pred <- SingleR(
       test = GetAssayData(all_seu, layer="data"),
       ref = ref,
       labels = ref$label.main
)

all_seu$SingleR_labels <- pred$labels

DimPlot(all_seu, group.by="SingleR_labels", label=TRUE)


```




